---
title: "Fruit and Vegetable Prices: EDLD651_final_project"
author: "Kyla Brannigan, Erika Davis, & Cheyna Warner"
date: "2025-12-10"
abstract: |
  insert abstract text here
abstract-title: "Fruit and Vegetable Prices"
format: 
  html:
    toc: true
    toc-location: left
    toc-title: Table of Contents
    code-fold: true
editor: visual
execute:
  eval: false
  echo: false
  warnings: false
  message: false
---
# Introduction

With the cost of living on the rise (SOURCE), it is important to determine how essential factors, such as food prices, are modulated. Though just a piece of the puzzle, the cost of fruits and vegetables can weigh heavily on consumers decisions in the grocery store (SOURCE). The central aim of this project is to understand how different factors impact retail price of fruits and vegetables in the United States. We selected factors such as yield, produce family, and produce form for the present analysis. 

The guiding questions for this study were: 
  1. How does the price of fruits and vegetables varies based on form?
  2. Do fruit and vegetable prices vary by family?
  3. What is the relationship between yield and retail price of produce? 

# Methods

For this project, the Fruit and Vegetable Prices data from the USDA Economic Research Service was used to shed light on questions such as how the price of fruits and vegetables varies based on form, if fruit and vegetable prices vary by family, and how yield impacts price of produce. This data contains information about over 150 types of fresh and processed fruit and vegetables.

To begin our analysis, we started by compiling and cleaning the raw data provided by the USDA Economic Research Service. To this end, we used the packages IDK HOW TO CITE THEM I FORGET.

```{r}
# Load Packages
library(here)
library(janitor)
library(rio)
library(tidyverse)
library(tm)
library(dplyr)
#install.packages("tm") <- i didn't have this package
```

First, we imported the data using the list.files argument and rbinded our files to make a fruit and a vegetable data frame. Once all of the data was imported, we had to clean and tidy the data frames as they were incredibly messy. In this step, we removed rows without data, renamed columns, and extracted irrelevant information such as citations and instructions. Code to complete this action can be found below.  
```{r}
# Import Data

# Fruit data

fruit_files <- list.files(pattern = ".xlsx", path = "../data/fruit-2020", full.names = TRUE)

fruit_2020 <- do.call(rbind, lapply
        (fruit_files, import, as.is=T, header = FALSE))

# Vegetable Data
veg_files <- list.files(pattern = ".xlsx", path = "../data/vegetables-2020", full.names = TRUE)

veg_2020 <- do.call(rbind, lapply
        (veg_files, import, as.is=T, header = FALSE))
```

```{r}
# Removing rows without data

# Fruit cleaning
fruit_2020 <- fruit_2020 %>% # rename cols to be more descriptive
  rename(fruit = ...1, retail_price = ...2, retail_price_unit = ...3, yield = ...4, cup_eq_size = ...5, cup_eq_unit = ...6,cup_eq_price = ...7)

fruit_2020_clean <- fruit_2020 %>% # remove rows without data (these were just the citation of where the data came from)
  filter(!substr(fruit, 1, 1) == "1") %>% 
  filter(!substr(fruit, 1, 1) == "2") %>% 
  filter(!substr(fruit, 1, 1) == "3") %>% 
  filter(!substr(fruit, 1, 1) == "4") %>% 
  filter(!substr(fruit, 1, 6) == "Source")

# Veggie cleaning
veg_2020 <- veg_2020 %>% # rename cols to be more descriptive
  rename(veg = ...1, retail_price = ...2, retail_price_unit = ...3, yield = ...4, cup_eq_size = ...5, cup_eq_unit = ...6,cup_eq_price = ...7)

veg_2020_clean <- veg_2020 %>% # remove rows without data (these were just the citation of where the data came from)
  filter(!substr(veg, 1, 1) == "1") %>% 
  filter(!substr(veg, 1, 1) == "2") %>% 
  filter(!substr(veg, 1, 1) == "3") %>% 
  filter(!substr(veg, 1, 1) == "4") %>% 
  filter(!substr(veg, 1, 6) == "Source")
```

```{r}
# Tidying the data

# Fruit
fruit_names <- c("Apples", "Apricots", "Bananas", "Mixed berries", "Blackberries", "Blueberries", "Cantaloupe", "Cherries", "Clementines", "Cranberries", "Dates", "Figs", "Fruit cocktail", "Grapefruit", "Grapes", "Honeydew melon", "Kiwi", "Mangoes", "Nectarines", "Oranges", "Papaya", "Peaches", "Pears", "Pineapple", "Plums", "Pomegranate", "Raspberries", "Strawberries", "Watermelon") # creating a vector with the fruit names

fruit_2020_clean <- fruit_2020_clean |> 
  # fix the fruit column to isolate the fruit name
  mutate(fruit = str_extract(fruit, "[^—]+")) |> 
  # create a flag for fruit names
  mutate(is_fruit = fruit %in% fruit_names) |> 
  # fill the fruit name down
  mutate(fruit_name = if_else(is_fruit, fruit, NA_character_)) |> 
  fill(fruit_name, .direction = "down") |> 
  # remove rows where fruit column is the fruit name or "Form"
  filter(!is_fruit, fruit != "Form") |> 
  # rename columns
  rename(form = fruit) |> 
  # move the fruit column to the beginning
  relocate(fruit_name, .before = "form") |> 
  # remove the is_fruit column, we don't need it anymore
  select(-is_fruit) |> 
  # remove any rows with NA in the retail_price column
  filter(!is.na(retail_price))

# Vegetable
veg_names <- c("Acorn squash", "Artichoke", "Asparagus", "Avocados", "Beets", "Black beans", "Blackeye peas", "Broccoli", "Brussels sprouts", "Butternut squash", "Cabbage", "Carrots", "Cauliflower", "Celery", "Collard greens", "Sweet corn", "Cucumbers", "Great northern beans", "Green beans", "Green peas", "Green peppers", "Kale", "Kidney beans", "Lentils", "Iceberg lettuce", "Romaine lettuce", "Lima beans", "Mixed vegetables", "Mushrooms", "Mustard greens", "Navy beans", "Okra", "Olives", "Onions", "Pinto beans", "Potatoes", "Pumpkin", "Radish", "Red peppers", "Spinach", "Sweet potatoes", "Tomatoes", "Turnip greens", "Zucchini") # creating a vector with the veggie names

veg_2020_clean <- veg_2020_clean |> 
  # fix the veg column to isolate the veg name
  mutate(veg = str_extract(veg, "[^—]+")) |> 
  # create a flag for veg names
  mutate(is_veg = veg %in% veg_names) |> 
  # fill the veg name down
  mutate(veg_name = if_else(is_veg, veg, NA_character_)) |> 
  fill(veg_name, .direction = "down") |> 
  # remove rows where veg column is the veg name or "Form"
  filter(!is_veg, veg != "Form") |> 
  # rename columns
  rename(form = veg) |> 
  # move the veg column to the beginning
  relocate(veg_name, .before = "form") |> 
  # remove the is_veg column, we don't need it anymore
  select(-is_veg) |> 
  # remove any rows with NA in the reatil_price column
  filter(!is.na(retail_price))
```

For our research questions, some specific data formatting and tidying was necessary. For research question one, how the price of fruits and vegetables varies based on form, we needed to alter the form category to better allow for analysis. For example, tomatoes were listed as "Grape", "Roma", and "Large" rather than just fresh. We simple converted the information to say "Fresh" so it was in alignment with other vegetables in the data frame. 

```{r}
# How does the price of fruits and vegetables vary based on form? (Kyla)

# remove the numbers at the end of the forms
fruit_2020_clean$form <- removeNumbers(fruit_2020_clean$form)
veg_2020_clean$form <- removeNumbers(veg_2020_clean$form)

# remove additional specifiers from form column that are not important
fruit_2020_clean$form <- gsub("\\s*\\([^\\)]+\\)", "", fruit_2020_clean$form)
fruit_2020_clean$form <- sub(",.*", "", fruit_2020_clean$form)
veg_2020_clean$form <- sub(" .*$", "", veg_2020_clean$form)
veg_2020_clean$form <- gsub(",", "", veg_2020_clean$form)

# change the tomato forms from "Grape", "Roma", and "Large to all be "Fresh"
# and filter to only look at price differences between fresh, frozen, canned, and dried
veg_2020_clean <- veg_2020_clean |> 
  mutate(form = case_when(form == "Canned" ~ "Canned",
                   form == "Grape" ~ "Fresh", 
                   form == "Roma" ~ "Fresh", 
                   form == "Large" ~ "Fresh",
                   TRUE ~ form),
         form = as_factor(form)) |> 
  filter(form == "Fresh" | form == "Frozen" | form == "Canned" | form == "Dried")

# remove the specific rows for raisins and applesauce because they don't fit cleanly into a category
# and change "Ready to drink" to "Juice"
fruit_2020_clean <- fruit_2020_clean |> 
  filter(form != "Raisins" & form != "Applesauce") |> 
  mutate(form = case_when(form == "Ready to drink" ~ "Juice",
                          TRUE ~ form),
         form = as_factor(form))

# make sure retail_price is a numeric variable
fruit_2020_clean <- fruit_2020_clean |> 
  mutate(retail_price = as.numeric(retail_price))
veg_2020_clean <- veg_2020_clean |> 
  mutate(retail_price = as.numeric(retail_price))

# examine retail price by form
fruit_summary <- fruit_2020_clean |> 
  mutate(form = as_factor(form)) |> 
  group_by(form) |> 
  summarize(mean_retail_price = mean(retail_price, na.rm = TRUE),
            sd_price = sd(retail_price, na.rm = TRUE),
            n = n(),
            se_price = sd_price / sqrt(n))
veg_summary <- veg_2020_clean |> 
  mutate(form = as_factor(form)) |> 
  group_by(form) |> 
  summarize(mean_retail_price = mean(retail_price, na.rm = TRUE),
            sd_price = sd(retail_price, na.rm = TRUE),
            n = n(),
            se_price = sd_price / sqrt(n))
```

Our second research question required the most synthesis as it pertained to information not originally provided in the data set. To supplement, we IDK ERIKA WHATEVER RESEARCH YOU DID.

```{r}
# Do fruit and vegetable prices vary by family (berries, stone fruit etc)? (Erika)

#making a new column for the fruit groups -- I can add where I found this info in our introduction :P

fruit_family_data <- fruit_2020_clean %>%
   mutate(
    family = case_when(
      fruit_name %in% c("Peaches", "Plums", "Apricots", "Cherries",
                        "Nectarines", "Mangoes", "Dates") ~ "stone fruit",
      
      fruit_name %in% c("Blackberries", "Blueberries", "Raspberries",
                        "Strawberries", "Pomegranate", "Bananas", "Kiwi",
                        "Pineapple", "Cranberries", "Grapes", "Papaya") ~ "berries",
      
      fruit_name %in% c("Apples", "Pears") ~ "pome",
      
      fruit_name %in% c("Clementines", "Oranges", "Grapefruit") ~ "citrus",
      
      fruit_name %in% c("Cantaloupe", "Honeydew melon", "Watermelon") ~ "melon",
      
      fruit_name %in% c("Mixed berries", "Fruit cocktail") ~ "mixed fruits",
      
      fruit_name %in% c("Figs") ~ "syconium",
      
      TRUE ~ "other"
    )
  )

# doing the same exact thing but for vegetables 

veg_family_data <- veg_2020_clean %>%
   mutate(
    family = case_when(
      veg_name %in% c("Broccoli", "Brussels sprouts", "Cabbage", "Cauliflower",
                        "Collard greens", "Kale", "Turnip greens", "Radish", "Mustard greens") ~ "cabbage family",
      
      veg_name %in% c("Carrots", "Celery") ~ "carrot family",
      
      veg_name %in% c("Onions", "Asparagus") ~ "liliceae family",
      
      veg_name %in% c("Acorn squash", "Cucumbers", "Butternut squash", "Pumpkin", "Zucchini") ~ "gourd family",
      
      veg_name %in% c("Black beans", "Blackeye peas", "Great northern beans", "Green beans", "Green peas", "Kidney beans", "Lima beans", "Navy beans", "Pinto beans", "Lentils") ~ "legume family",
      
      veg_name %in% c("Tomatoes", "Sweet potatoes", "Red peppers", "Potatoes", "Green peppers") ~ "nightshade family",
      
      veg_name %in% c("Okra") ~ "hibiscus family",
      
      veg_name %in% c("Sweet corn") ~ "grass family",
      
      veg_name %in% c("Artichoke", "Iceberg lettuce", "Romaine lettuce") ~ "aster family",
      
      veg_name %in% c("Spinach", "Beets") ~ "goosefoot family",
      
      veg_name %in% c("Avocados", "Olives") ~ "botanically considered a fruit",
      
      veg_name %in% c("Mushrooms") ~ "fungus",
      
      veg_name %in% c("Mixed vegetables") ~ "mixed",
      
      TRUE ~ "other"
    )
  )

str(veg_family_data$cup_eq_price)
veg_family_data <- veg_family_data %>% 
  mutate(cup_eq_price = as.numeric(cup_eq_price))

```

Finally, for research question three minimal alterations were made to the data frame for analysis. A column indicating if a produce item was a fruit or a vegetable was added to assist in grouping for analysis. 

```{r}
# CHEYNA
# examine the relationship between yield and price
fruit_2020_clean <- fruit_2020_clean %>% 
  rename(name = fruit_name) %>% #rename col to join fruit df with veg df
  filter(form == "Fresh"| form == "Frozen" | form == "Canned" |form == "Dried") # select forms common across produce 
  fruit_2020_clean$fruit_veg <- "Fruit" 

veg_2020_clean <- veg_2020_clean %>% 
  rename(name = veg_name)
  veg_2020_clean$fruit_veg <- "Vegetable" 

# make fruit and veg df
fruit_veg_df <- bind_rows(fruit_2020_clean, veg_2020_clean)
```

# Results

Below is a descriptive table of the data. Results are presented by research question. 

```{r}
# KYLA
# ANOVAs to look at how retail price varies by fruit and veggie form
fruit_form_anova_result <- aov(retail_price ~ form, data = fruit_2020_clean)
summary(fruit_form_anova_result)
pairwise.t.test(fruit_2020_clean$retail_price, fruit_2020_clean$form,
                p.adjust.method = "bonferroni")

veggie_form_anova_result <- aov(retail_price ~ form, data = veg_2020_clean)
summary(veggie_form_anova_result)

# plotting average retail price by fruit form
fruit_form_plot <- fruit_summary |> 
  ggplot(aes(x = form, y = mean_retail_price, fill = form)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_retail_price - se_price, ymax = mean_retail_price + se_price), width = 0.3) +
  labs(title = "Average Fruit Retail Prices by Form",
       x = "Form",
       y = "Average Retail Price") +
  scale_fill_manual(values = c("#005f73", "#0a9396", "#94d2bd", "#e9d8a6", "#ee9b00", "#ca6702", "#bb3e03")) +
  theme(theme_minimal(),
        legend.position = "NULL",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_blank(),  
        plot.background = element_blank())

# plotting average retail price by veggie form
veg_form_plot <- veg_summary |> 
  ggplot(aes(x = form, y = mean_retail_price, fill = form)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_retail_price - se_price, ymax = mean_retail_price + se_price), width = 0.3) +
  labs(title = "Average Vegetable Retail Prices by Form",
       x = "Form",
       y = "Average Retail Price") +
  scale_fill_manual(values = c("#005f73", "#0a9396", "#94d2bd", "#e9d8a6", "#ee9b00", "#ca6702", "#bb3e03")) +
  theme(theme_minimal(),
        legend.position = "NULL",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_blank(),  
        plot.background = element_blank())
```

The retail price of fruits varied significantly by form, (F(6) = 12.59, p < .001).In running a post-hoc test, dried fruits were priced significantly higher than all other forms. However, there were no other differences between forms. The retail price of vegetables did NOT vary significantly by form (F(3) = 1.424, p = .243). 
 
```{r}
# ERIKA 
#looking at summary stats by family ... using cup eq. price so we have a relatively similar amount by price
str(fruit_family_data$cup_eq_price)
fruit_family_data <- fruit_family_data %>% 
  mutate(cup_eq_price = as.numeric(cup_eq_price))

fruit_family_data %>%
  group_by(family) %>%
  summarize(
    mean_price = mean(cup_eq_price, na.rm = TRUE),
    median_price = median(cup_eq_price, na.rm = TRUE),
    sd_price = sd(cup_eq_price, na.rm = TRUE),
    n = n()
  )

# running a one way anova to see if price (continuous) varies by family (categorical). I found that it does not vary signifiantly (p = .158)
anova_result <- aov(cup_eq_price ~ family, data = fruit_family_data)
summary(anova_result)

#plot
plot_table1 <- fruit_family_data %>%
  group_by(family) %>%
  summarize(
    mean_price = mean(cup_eq_price, na.rm = TRUE),
    sd_price = sd(cup_eq_price, na.rm = TRUE),
    n = n(),
    se_price = sd_price / sqrt(n)
  )

ggplot(plot_table1, aes(x = reorder(family, mean_price), y = mean_price)) +
  geom_col(fill = "deeppink4") +
  geom_errorbar(aes(ymin = mean_price - se_price, ymax = mean_price + se_price), width = 0.3) +
  labs(x = "Fruit Family", y = "Mean Price per Unit", title = "Mean Fruit Prices by Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# same thing but for vegetables 

anova_result2 <- aov(cup_eq_price ~ family, data = veg_family_data)
summary(anova_result2) # this one DID have a significant difference (p < 0.001) so i did post-hoc to see what's going on. 

pairwise.t.test(veg_family_data$cup_eq_price, veg_family_data$family,
                p.adjust.method = "bonferroni")

#confusing to read but the significant groups are: liliceae is signnificantly different from aster, liliceae different from botanical fruits, liliceae different from gourd, liliceae different from carrot, liliceae different from legume, liliceae different from mixed, liliceae different from nightshade
#difference mostly because of liliceae (onions + asparagus) being higher in price than others 

# plot 
summary_table <- veg_family_data %>%
  group_by(family) %>%
  summarize(
    mean_price = mean(cup_eq_price, na.rm = TRUE),
    sd_price = sd(cup_eq_price, na.rm = TRUE),
    n = n(),
    se_price = sd_price / sqrt(n)
  )

ggplot(summary_table, aes(x = reorder(family, mean_price), y = mean_price)) +
  geom_col(fill = "olivedrab4") +
  geom_errorbar(aes(ymin = mean_price - se_price, ymax = mean_price + se_price), width = 0.3) +
  labs(x = "Vegetable Family", y = "Mean Price per Unit", title = "Mean Vegetable Prices by Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Vegetables did vary significantly by family, F(12) = 4.786, p < .001. In running a post-hoc test, the Liliceae family (e.g. onions & asparagus) were priced significantly higher than all other families. There were no other differences between families. Fruits did not vary significantly by family, F(6) = 1.623, p = .158. 

```{r}
# CHEYNA 
# plot fruit and veg diff. 
fruit_veg_df <- fruit_veg_df %>% 
  mutate(yield = as.numeric(yield))

sorted_yield_price_plot <- fruit_veg_df %>% 
  filter(yield <= 2) %>% 
  ggplot(aes(x = yield, y = retail_price, color = fruit_veg)) +
  geom_point() +
  geom_smooth(method = "lm", fill = "lightgray") +
  scale_color_manual(values = c("#94d2bd", "#ee9b00")) +
  theme_minimal() +
  labs(title = "Relationship between Yield and Retail Price",
       x = "Yield", 
       y = "Retail Price",
       color = "Type")

sorted_yield_price_plot

# analysis 
sorted_yield_price_anova <- aov(retail_price ~ yield + fruit_veg, data = fruit_veg_df)

summary(sorted_yield_price_anova)
```

Finally, there was a significant difference between yield and price of fruits and yield and price of vegetables (F(1) = 16.991, p < .001). 
 
# Discussion 

# Conclusion 

# References 
